---
title: "DRAFT ROV/diver models"
author: "Rachael Aber"
date: "07/22/2026"
output: html_notebook
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE,
                      echo = T, warning = T, message = T,
                      fig.align = "center", fig.width = 6, fig.height = 3)

library(tidyverse)
library(lme4)
library(glmmTMB)
```

Read combined data. Data assembly (joining ROV + diver, category crosswalks,
0-100 -> 0-1 conversion, transect_id / key derivation) now lives upstream in
build_combined_abundance.R and build_combined_percent_cover.R -- this notebook
starts from their outputs, so it only ever contains modeling.
```{r}
dat_abundance <- read.csv("../results/combined/ROV_diver_abundance_combined.csv")
dat_pc <- read.csv("../results/combined/ROV_diver_percent_cover_combined.csv")
```

Percent-cover models (logistic example)
```{r}
dat_pc <- dat_pc %>%
  mutate(cover_red_algae_ind = if_else(cover_red_algae > 0, 1, 0),
         combined_green_algae_ind = if_else(combined_green_algae > 0, 1, 0),
         cover_crustose_coralline_ind = if_else(cover_crustose_coralline > 0, 1, 0),
         combined_substrate_boulder_ind = if_else(combined_substrate_boulder > 0, 1, 0),
         substrate_rock_.15.25cm.wa._ind = if_else(substrate_rock_.15.25cm.wa. > 0, 1, 0),
         combined_substrate_pebble_ind = if_else(combined_substrate_pebble > 0, 1, 0),
         substrate_sand_ind = if_else(substrate_sand > 0, 1, 0),
         substrate_shell_hash_ind = if_else(substrate_shell_hash > 0 , 1, 0)
         )

mod_ssh <- glmer(substrate_shell_hash_ind ~ type + site + season + depth + (1|transect_id),
                 data = dat_pc,
               family = binomial)

summary(mod_ssh)
```

Percent-cover models (binomial examples)
```{r}
# Note that it may be better to compute the numerators directly from detections out of n
dat_pc <- dat_pc %>%
  mutate(cover_red_algae_num = ceiling(cover_red_algae * n),
         combined_green_algae_num = ceiling(combined_green_algae * n),
         cover_crustose_coralline_num = ceiling(cover_crustose_coralline * n),
         combined_substrate_boulder_num = ceiling(combined_substrate_boulder * n),
         substrate_rock_.15.25cm.wa._num = ceiling(substrate_rock_.15.25cm.wa. * n),
         combined_substrate_pebble_num = ceiling(combined_substrate_pebble * n),
         substrate_sand_num = ceiling(substrate_sand * n),
         substrate_shell_hash_num = ceiling(substrate_shell_hash * n)
         )

mod_cra <- glmmTMB(
  cbind(cover_red_algae_num, n - cover_red_algae_num) ~ type + site + season + depth + (1|transect_id),
  family = binomial,
  data = dat_pc
)

summary(mod_cra)

mod_cra_disp <- glmmTMB(
  cbind(cover_red_algae_num, n - cover_red_algae_num) ~ type + site + season + depth + (1|transect_id),
  family = betabinomial(link = "logit"),
  data = dat_pc
)
summary(mod_cra_disp) # preferred

mod_cga_disp <- glmmTMB(
  cbind(combined_green_algae_num, n - combined_green_algae_num) ~ type + site + season + depth + (1|transect_id),
  family = betabinomial(link = "logit"),
  data = dat_pc
)
summary(mod_cga_disp)
```

Abundance visualization:
```{r}
ggplot(dat_abundance, aes(x = transect_id, y = ochre_mottled_star, col = type)) +
  geom_point() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

ggplot(dat_abundance, aes(x = transect_id, y = cancer_crab, col = type)) +
  geom_point() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

ggplot(dat_abundance, aes(x = transect_id, y = kelp_crab, col = type)) +
  geom_point()+
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

Abundance models: ochre mottled star example
```{r}
ochre_mod <- glmer(ochre_mottled_star ~ type + site + season + depth + (1|transect_id),
                   data = dat_abundance,
               family = poisson(link = "log"))
summary(ochre_mod)

# Check nb
ochre_mod_disp <- glmer.nb(ochre_mottled_star ~ type + site + season + depth + (1|transect_id),
                   data = dat_abundance)
summary(ochre_mod_disp)
```
